### 2025-02-26 paper/relatedwork.tex
You only linearize once.
Remove things definitely handled.
Note Etch.

% - "Verified Tensor-Program Optimization Via High-level Scheduling Rewrites"
% - "Efficient Differentiable Programming in a Functional Array-Processing Language"
% - "Verified tensor-program optimization via high-level scheduling rewrites"
% - "Indexed Streams: A Formal Intermediate Representation for Fused Contraction Programs"
% - "You only linearize once"
% - Dependent ML
% - ATS
% - Jax
% - Remora

@@

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
\todo[inline]{reviewer 3 (2024) notices: This paper does not really answer this question besides enforcing consistency of shapes and accesses? Besides, final extraction could break these properties?  this is correct, we have to clarify this.}
(ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.

@@

>>>>

@@

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
\todo[inline]{reviewer 3 (2024) notices: This paper does not really answer this question besides enforcing consistency of shapes and accesses? Besides, final extraction could break these properties?  this is correct, we have to clarify this.}
(ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.

@@

Another related approach is the Etch compiler for indexed
streams~\cite{10.1145/3591268}, which is implemented as an embedded
language in Lean. Although the domain differs from ours, stream/tensor
contraction as compared to array programming, the basic idea of
extracting low level code is similar. Significant differences include
our support for rank polymorphism, automatic differentiation, and
parallel execution.
